An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-22 article/article.tex
finished revision until 2.1

%% - \Abstract{} almost as usual
\Abstract{Performing inference in statistical models with intractable
  likelihood is challenging; most likelihood-free inference (LFI)
  methods encounter accuracy and efficiency limitations. In this
  paper, we present the implementation of Robust Optimisation Monte
  Carlo (ROMC) in the \proglang{Python} package \pkg{ELFI}. ROMC is a
  novel and efficient (highly-parallelisable) LFI framework that
  provides accurate weighted samples from the posterior. Our
  implementation can be used in two ways. First, a scientist may use
  it as an out-of-the-box LFI algorithm; we provide an easy-to-use API
  harmonised with the principles of \pkg{ELFI}, enabling effortless
  comparisons with the rest of the methods included in the
  package. Additionally, we have carefully split ROMC into isolated
  components for supporting extensibility. A researcher may experiment
  with novel method(s) for solving part(s) of ROMC without
  reimplementing everything from scratch. In both scenarios, all steps
  run in a fully-parallelisable manner, exploiting all CPU
  cores. We also provide helpful functionalities for (i) inspecting
  the inference process and (ii) evaluating the obtained
  samples. Finally, we test the robustness of our implementation on
  some typical LFI examples.  }

@@

%% -- Introduction -------------------------------------------------------------
\section{Introduction} \label{sec:intro}
\input{latex_files/ch1/1-introduction.tex}

% \clearpage
\section{Background}
\input{latex_files/ch2/2-0-intro.tex}

\subsection{Simulator-based models and likelihood-free inference}
\input{latex_files/ch2/2-1-implicit-models.tex}

\subsection{Optimisation Monte Carlo (OMC)}
\input{latex_files/ch2/2-2-OMC.tex}

\subsection{Robust optimisation Monte Carlo (ROMC)}
\input{latex_files/ch2/2-3-ROMC.tex}

@@

%% -- Design Principles -----------------------------------------------------
\section{Implementation}
This section is split in two parts. We first express ROMC as an
algorithm and then we present the general implementation principles we
follow.

\subsection{Algotirhmic view of ROMC}
\input{latex_files/ch3/3-0-design-principles.tex}

>>>>

%% - \Abstract{} almost as usual
\Abstract{Performing inference in statistical models with an
  intractable likelihood is challenging, therefore, most
  likelihood-free inference (LFI) methods encounter accuracy and
  efficiency limitations. In this paper, we present the implementation
  of the LFI method Robust Optimisation Monte Carlo (ROMC) in the
  \proglang{Python} package \pkg{ELFI}. ROMC is a novel and efficient
  (highly-parallelisable) LFI framework that provides accurate
  weighted samples from the posterior. Our implementation can be used
  in two ways. First, a scientist may use it as an out-of-the-box LFI
  algorithm; we provide an easy-to-use API harmonised with the
  principles of \pkg{ELFI}, enabling effortless comparisons with the
  rest of the methods included in the package. Additionally, we have
  carefully split ROMC into isolated components for supporting
  extensibility. A researcher may experiment with novel method(s) for
  solving part(s) of ROMC without reimplementing everything from
  scratch. In both scenarios, the ROMC parts can run in a
  fully-parallelised manner, exploiting all CPU cores. We also provide
  helpful functionalities for (i) inspecting the inference process and
  (ii) evaluating the obtained samples. Finally, we test the
  robustness of our implementation on some typical LFI examples.}

@@

%% -- Introduction -------------------------------------------------------------
\section{Introduction} \label{sec:intro}

Simulator-based models are particularly captivating due to the
provided modeling freedom. In essence, any data generating mechanism
that can be written as a finite set of algorithmic steps can be
programmed as a simulator-based model. In these cases, it is feasible
to generate samples using the simulator but it is infeasible to
evaluate the likelihood function. The intractability of the likelihood
makes the likelihood-free inference (LFI), i.e., the approximation of
the posterior distribution without using the likelihood function,
particularly challenging.

Optimization Monte Carlo (OMC) proposed by~\citet{Meeds2015} is a
novel LFI approach for approximating the posterior distribution. The
central idea is turning the stochastic data-generating mechanism into
a set of deterministic optimization processes. Afterwards,
\citet{Forneron2016} provided a similar method under the name `reverse
sampler'. In their work,~\citet{Ikonomov2019}, located some critical
limitations of OMC, so they proposed Robust OMC (ROMC), an alternative
version of OMC with the appropriate modifications.

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (Engine for likelihood-free
inference)~\citet{1708.00707}. As we illustrate at
Section~\ref{sec:implementation}, we have carefully designed the
implementation to ensure extensibility. ROMC is a general framework
for obtaining weighted samples from the posterior, i.e., it defines a
sequence of algorithmic steps without enforcing a specific algorithm
for solving each step. Therefore, a researcher may use ROMC as the
backbone algorithm and develop novel methods to solve each separate
step.\footnote{For being a ready-to-use
algorithm,~\citet{Ikonomov2019} proposed a default method for each
step, but this choice is by no means restrictive.} We have designed
our software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
evaluation in three steps. First, for securing that our implementation
is accurate, we test it against an artificial example with a tractable
likelihood. The artificial example also serves as a step-by-step guide
for showcasing how to use the various functionalities of our
implementation. Second, we use the second-order moving average (MA2)
example from the \pkg{ELFI} package, using as ground truth the samples
obtained with Rejection ABC~\citet{lintusaari2017}, with a very large
number of trials. Finally, we present the execution times of ROMC,
measuring the speed-up obtained by using the parallel version of the
implementation.

% \clearpage
\section{Background}

We first give a short introduction to simulator-based models, we then
focus on OMC and its robust improvement, ROMC, and we, finally,
introduce \pkg{ELFI}, the underlying package that is used for the
implementation of ROMC.

\subsection{Simulator-based models and likelihood-free inference}

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism. The key characteristic of these models is
that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterized
family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these scenarios, we can only a
access the simulator \( m_r(\thetab) \), i.e., a black-box mechanism
(computer code) that generates samples \(\yb\) in a stochastic manner
from a set of parameters \(\thetab\). We denote the process of
obtaining samples from the simulator with
\( m_r(\thetab) \rightarrow \yb \). It is feasible to isolate the
randomness of the simulator by introducing the nuisance random
variables denoted by \(\ub \sim p(\ub)\). Therefore, for a tuple
\((\thetab, \ub)\), the simulator turns to a deterministic mapping
\(g\), such that \(\yb=\simulator(\thetab,\ub)\). In terms of computer
code, the randomness of a random process is governed by the global
seed. Even though each software may handle the random generation
process in a different way\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, by sampling an integer value
and setting it as the initial seed converts the simulation to a
deterministic sequence of steps.

Simulator-based models provide considerable modeling freedom; any
physical process that can be conceptualized as a computer program of
finite steps can be modeled as a simulator-based model without any
compromise. The modeling freedom allows for any amount of hidden
(unobserved) internal variables or rule-based decisions. Hence,
implicit models are often used to model physical phenomena in the
natural sciences such as, e.g., genetics, epidemiology or
neuroscience. Further background on simulator-based models and
example applications can be found in the articles
by~\citet{gutmann2016, lintusaari2017, sisson2018, cranmer2020}.

The modeling freedom of simulator-based models, however, comes at the
price of difficulties in inferring their parameters. Denoting the
observed data as \(\data\), the main difficulty is that the likelihood
function \(l(\thetab) = p(\data|\thetab)\) is generally
intractable. To better see the sources of the intractability, and to
address them, we go back to the basic characterization of the
likelihood as the (rescaled) probability of a parameter of the model
to generate data \(\yb\) that is similar to the observed data
\(\data\). More formally, the likelihood \(l(\thetab)\) equals
\begin{equation} \label{eq:likelihood}
  l(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in b_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around \(\data\)
that is defined via a distance function \(d\), i.e., \ \(\region(\data)
:= \{\yb: d(\yb, \data) \leq \epsilon \}\).

The basic characterization of the likelihood in~\eqref{eq:likelihood}
highlights two sources of intractability; the first is the computation
of the probability
\(\Pr(\simulator(\thetab,\ub) \in \region(\data))\), the second is the
limit of \(\epsilon \to 0\). Approximating the probability with
samples becomes computationally infeasible if \(\epsilon\) is too
small. Hence, a large class of inference methods work with
\(\epsilon > 0\), which leads to the approximate likelihood function
\(l_{d, \epsilon}(\thetab)\)
\begin{equation} \label{eq:approx_likelihood}
  l_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

and, in turn, to the approximate posterior

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data) \propto l_{d, \epsilon}(\thetab) p(\thetab).
\end{equation}

The approximation in~\eqref{eq:approx_likelihood} is by no means the
only strategy to deal with the intractability of the likelihood
function in~\eqref{eq:likelihood}. Other strategies include modeling
the (stochastic) relationship between \(\thetab\) and \(\yb\), and its
reverse, or framing likelihood-free inference as a ratio estimation
problem, see for example \citet{blum2010, Wood2006, Papamakarios2016,
  Papamakarios2019, Chen2019, Thomas2020, Hermans2020}. However, both
OMC and robust OMC, which we introduce next, are based on the
approximation in~\eqref{eq:approx_likelihood}.

\subsection{Optimization Monte Carlo (OMC)}
\input{latex_files/ch2/2-2-OMC.tex}

\subsection{Robust optimization Monte Carlo (ROMC)}
\input{latex_files/ch2/2-3-ROMC.tex}

@@

%% -- Design Principles -----------------------------------------------------
\section{Implementation}
\label{sec:implementation}
This section is split in two parts. We first express ROMC as an
algorithm and then we present the general implementation principles we
follow.

\subsection{Algorithmic view of ROMC}
\input{latex_files/ch3/3-0-design-principles.tex}

### 2022-12-22 article/latex_files/ch1/1-introduction.tex
fixed spaces and commas
finished revision until 2.1

Simulator-based models are particularly captivating due to the
modelling freedom they provide. In essence, a simulator-based model
can describe any data generating mechanism that can be written as a
finite set of algorithmic steps. Such freedom comes at a cost;
performing the inference, i.e., sampling or evaluating the posterior,
is challenging.

Optimisation Monte Carlo (OMC) proposed by~\citet{Meeds2015} is a
novel LFI approach for approximating the posterior distribution. The
central idea is turning the stochastic data-generating mechanism into
a set of deterministic optimisation processes. Afterwards,
\citet{Forneron2016} provided a similar method under the name `reverse
sampler'. In their work, ~\citet{Ikonomov2019}, located some critical
limitations of OMC. They proposed Robust OMC (ROMC), an alternative
version of OMC with appropriate modifications.

In this paper, we present an implementation of ROMC at the
\proglang{Python} package \pkg{Engine for Likelihood-Free inference
(ELFI)}. We follow appropriate design principles for ensuring
extensibility. As we describe analytically in the next chapter, ROMC
is a general framework for obtaining weighted samples from the
posterior; it defines a sequence of algorithmic steps without
enforcing a specific algorithm for solving each step\footnote{For
being a ready-to-use algorithm,~\citet{Ikonomov2019} proposed a
default method for each step, but this choice is by no means
restrictive.}. A researcher may adopt ROMC and develop novel methods
to solve each task. We have designed our software for facilitating
such alterations. Finally, we have tested the accuracy and the
efficiency of our implementation on some LFI examples supported by the
\pkg{ELFI} package.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework, so there are no
other packages for direct comparisons. Therefore, we test ROMC against
(i) an artificial example with tractable likelihood and (ii) the
second-order moving average (MA2) example from the \pkg{ELFI}
package. In the latter, we generate data using known parameters and
compare our results with the typical Rejection
ABC~\citet{lintusaari2017}.

>>>>

Simulator-based models are particularly captivating due to the
provided modeling freedom. In essence, any data generating mechanism
that can be written as a finite set of algorithmic steps can be
programmed as a simulator-based model. In these cases, it is feasible
to generate samples using the simulator but it is infeasible to
evaluate the likelihood function. The intractability of the likelihood
makes the likelihood-free inference (LFI), i.e., the approximation of
the posterior distribution without using the likelihood function,
particularly challenging.

Optimization Monte Carlo (OMC) proposed by~\citet{Meeds2015} is a
novel LFI approach for approximating the posterior distribution. The
central idea is turning the stochastic data-generating mechanism into
a set of deterministic optimization processes. Afterwards,
\citet{Forneron2016} provided a similar method under the name `reverse
sampler'. In their work,~\citet{Ikonomov2019}, located some critical
limitations of OMC, so they proposed Robust OMC (ROMC), an alternative
version of OMC with the appropriate modifications.

In this paper, we present the implementation of ROMC at the
\proglang{Python} package \pkg{ELFI} (Engine for likelihood-free
inference)~\citet{1708.00707}. As we illustrate at
Section~\ref{sec:implementation}, we have carefully designed the
implementation to ensure extensibility. ROMC is a general framework
for obtaining weighted samples from the posterior, i.e., it defines a
sequence of algorithmic steps without enforcing a specific algorithm
for solving each step. Therefore, a researcher may use ROMC as the
backbone algorithm and develop novel methods to solve each separate
step.\footnote{For being a ready-to-use
algorithm,~\citet{Ikonomov2019} proposed a default method for each
step, but this choice is by no means restrictive.} We have designed
our software for facilitating such experimentation.

To the best of our knowledge, this is the first implementation of the
ROMC inference method to a generic LFI framework. We organize the
evaluation in three steps. First, for securing that our implementation
is accurate, we test it against an artificial example with a tractable
likelihood. The artificial example also serves as a step-by-step guide
for showcasing how to use the various functionalities of our
implementation. Second, we use the second-order moving average (MA2)
example from the \pkg{ELFI} package, using as ground truth the samples
obtained with Rejection ABC~\citet{lintusaari2017}, with a very large
number of trials. Finally, we present the execution times of ROMC,
measuring the speed-up obtained by using the parallel version of the
implementation.
